---
title: "Overwatch League Season 2022"
author: "Israel Garcia C."
date: "`r Sys.Date()`"
output: 
 html_document:
  theme:
    #bootswatch: darkly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(DT)
library(ggimage)
library(lubridate)
library(tidyverse)
owl_data_2022_scores <- read_rds("Data_all\\owl_data_2022_scores.rds")
owl_images <- read_rds("Data_all\\owl_logos.rds")
```

## Overwatch League Season 2022 Analysis
Thanks to the [official page](https://overwatchleague.com/en-us/statslab) of the league, people can search and play with the stats
produced in each match. Although it provides an easy way to get insight, I found it difficult to use for specific hypothesis or topics that people might want to know.
Also, it includes information of previous seasons, but they have a **different format** both in the tournaments and in the
number of players, **different map** pools, between other aspects. That makes the analysis of the current season with previous information kind of useless.  
Some general considerations:  
1. The last update of the official dataset was in `10/26/2022`, but the season has already finish.\
 - 1.1 This makes that the analysis incomplete, missing part of the information for the `Postseason` and all information of the `Playoffs`.\
2. This version might not be as interactive as in the official page, I plan to make it a `Shiny App` later or maybe for the following season. Another option was to make it this analysis using `SQL` and `Power BI/Tableau`, but for practice I choose to make it all in R and in Markdown.\
3. Graphics need a correct theme, I know but I am doing this in my free time as data analyst. I am trying to improve analysis and visualization. Also, I know the technical aspect of R Markdown, but I lack the ability of making pretty documents.

### Preview
As always, it is useful for the **user** and the developer to have a preview of the data in the analysis.  
Although the data in the official dataset is *almost* clean, there are some data entry errors and missing useful columns.  
Here, you can see a clean preview, first 5 entries, with additional information that I have added.
```{r Preview, echo = FALSE, results='asis'}
datatable(head(owl_data_2022_scores, 20), 
          options = list(
          columnDefs = list(list(className = 'dt-center', targets = 5)),
          pageLength = 5,
          lengthMenu = c(5, 10, 15, 20),
          extensions = 'FixedColumns',
          dom = 't',
          scrollX = TRUE,
          fixedColumns = TRUE))
```


### Match duration
The first part is to look into the match duration, in hours, for all the matches.  
As expected, *Tournaments* tend to have a larger duration than Qualifiers.
This can be explained by factors such as when you lose in a tournament you are knockoff so teams try their best to don't get eliminated as quick as possible. Also, the final match of tournaments is play in a *first to 4* format and not *first to 3*.
```{r Distribution of the matches duration in Hours, echo=FALSE, warning = FALSE}
# Distribution of the matches duration in Hours ----
owl_data_2022_scores %>% 
  select(match_id, game_number, stage_class, total_time_match) %>% 
  group_by(match_id) %>% 
  filter(row_number() == 1, stage_class != "Postseason") %>%
  ggplot(aes(x = as.numeric(as.duration(total_time_match), "hour"))) +
  geom_histogram(binwidth = 0.1) +
  facet_grid(stage_class ~ ., scales = "free_y") + 
  labs(x = "Total match time (Hours)",
       y = "Total matches",
       title = "OWL Season 2022 Match duration",
       caption = "Source: https://overwatchleague.com")

# Box Plot distribution of rounds. ----
owl_data_2022_scores %>%  
  select(match_id, game_number, match_rounds, stage_class, total_time_match) %>% 
  group_by(match_id) %>% 
  filter(game_number == max(game_number), match_rounds == max(match_rounds), stage_class != "Postseason") %>%
  ggplot(aes(x = as.numeric(as.duration(total_time_match), "hour"), y = match_rounds)) +
  geom_boxplot() +
  coord_flip() +
  facet_grid(stage_class ~ ., scales = "free_y") +
  labs(x = "Total match time (Hours)",
       y = "Total rounds per match",
       title = "OWL Season 2022 Match duration",
       caption = "Source: https://overwatchleague.com")
```

### Map pool
Overwatch counts with a variety of maps, not all of them are eligible during certain qualifiers or tournaments.  
In general, the first team to win 3 maps win the match. The order of the maps according to their type is:   
1. Control map  
2. Escort map  
3. Hybrid map, at this point it is *expected* that a team has won by a score of *3-0*. (only in Qualifiers)  
4. Push map  
5. Another control map  
On **Qualifiers**, we have the same number of Hybrid and Escort maps as they are played only once per match. Push is the fourth map to get played if there is no winner in the first three maps, and another map of Control is played as a tie breaker.  
On **Tournaments**, we have less teams and the final match is play in a *4-0* format.
```{r Map pool, echo=FALSE, warning = FALSE}
owl_data_2022_scores %>%
  select(match_id, game_number, match_winner, map_name, total_time_match, stage_class, map_class) %>% 
  group_by(match_id, game_number) %>% 
  filter(row_number() == 1, stage_class != "Postseason") %>% 
  ggplot(aes(x = map_class, fill = map_class)) +
  geom_bar() +
  geom_text(aes(label = ..count..), stat = "count", hjust = -0.1) +
  coord_flip() +
  facet_grid(stage_class ~ .) +
  labs(x = NULL,
       y = "Total times played",
       title = "OWL Season 2022",
       subtitle = "Maps played during this season",
       caption = "Source: https://overwatchleague.com",
       fill = "Map type") +
  paletteer::scale_fill_paletteer_d("rtist::vangogh")
```

### Scores in Overwatch League
As mentioned previously:  
Qualifiers are play as a **first to 3** format.  
In Qualifiers, the maximum is 3-2 as in case of tie, 2-2, the last map is Push, where it is "impossible" to draw again.  
Tournaments Qualifiers are play as a **first to 4**.  
```{r Common scores, echo=FALSE}
owl_data_2022_scores_common <- owl_data_2022_scores %>% 
  select(match_id, score_winner, score_loser, stage_class) %>% 
  group_by(match_id) %>% 
  filter(row_number() == 1) %>%
  mutate(score_str = str_c(score_winner, score_loser, sep = "-"))

owl_data_2022_scores_common %>%
  filter(stage_class != "Postseason") %>%
  ggplot(aes(x = score_str, fill = score_str)) +
  geom_bar() +
  coord_flip() +
  facet_wrap(. ~ stage_class, nrow = 2) +
  labs(x = NULL,
       y = "Total matches",
       title = "Common scores during season 2022",
       caption = "Source: https://overwatchleague.com",
       fill = "Final Score") +
  paletteer::scale_fill_paletteer_d("rtist::vangogh")
```

### Maps taken
As in any sports, both teams fight to get points to win. In Overwatch each point is attributed by winning a map. Obviously, the number of maps taken by losers will be less than the winners.
```{r Maps taken, echo = FALSE, message=FALSE}
owl_data_2022_scores %>%
  select(match_id, score_winner, score_loser) %>% 
  group_by(match_id) %>% 
  filter(row_number() == 1) %>%
  ungroup() %>% 
  summarise(`Maps taken by winning team` = sum(score_winner),
            `Maps taken by loser team` = sum(score_loser))

owl_data_2022_scores_maps_taken <- owl_data_2022_scores %>% 
  select(match_id, game_number, map_class, map_winner, match_winner, stage_class) %>% 
  group_by(match_id, game_number) %>% 
  filter(row_number() == 1)

owl_data_2022_scores_maps_taken %>% 
  group_by(map_class) %>% 
  summarise(wins = ifelse((map_winner == match_winner), 1 , 0)) %>%
  mutate(wins = ifelse(wins == 1,"Winner", "Loser")) %>% 
  count(wins) %>% 
  ggplot(aes(x = map_class, y = n, fill = wins)) +
  geom_col(position = "dodge") +
  geom_text(aes(label = n), vjust = -0.5, position = position_dodge(0.9)) +
  labs(x = NULL,
       y = "Total maps",
       title = "Maps taken by winner/loser match team",
       caption = "Source: https://overwatchleague.com",
       fill = "Maps taken by:") +
  paletteer::scale_fill_paletteer_d("rtist::vangogh")

owl_data_2022_scores_maps_taken %>% 
  group_by(map_class, stage_class) %>% 
  filter(stage_class != "Postseason") %>% 
  summarise(wins = ifelse((map_winner == match_winner), 1 , 0)) %>%
  count(wins) %>% 
  mutate(wins = ifelse(wins == 1,"Winner", "Loser")) %>% 
  ggplot(aes(x = map_class, y = n)) +
  geom_bar(aes(fill = wins), stat = "identity", position = "dodge") +
  coord_flip() + 
  facet_wrap(. ~ stage_class, nrow = 2) +
  labs(x = NULL,
       y = "Total maps",
       title = "Maps taken by the team",
       caption = "Source: https://overwatchleague.com",
       fill = "Map taken by:") +
  paletteer::scale_fill_paletteer_d("rtist::vangogh")
```

### Reverse sweep
One of the most entertaining/frustrating moments in any sports are the reverse sweeps.\
In Overwatch a reverse sweep happens when a team win all consecutive maps except for their match point map lose the following maps in a row getting a defeat.  
```{r reversed sweep, echo=FALSE}
owl_data_2022_scores %>%
  select(match_id, game_number, stage_class, map_winner, match_winner) %>% 
  group_by(match_id, game_number) %>% 
  filter(row_number() == 1, stage_class != "Postseason") %>%
  group_by(match_id) %>% 
  filter(game_number <= 2) %>%
  mutate(winner_second_map = lag(map_winner)) %>% 
  filter(game_number == 2) %>% 
  mutate(reversed_possible = ifelse(map_winner == winner_second_map, 1, 0),
         reversed = ifelse(reversed_possible == 1 & match_winner != map_winner,
                           "Reversed", "No")) %>% 
  ungroup() %>% 
  count(reversed)
```

From the numbers, it seems that a reverse sweep is quite infrequent, just `18 matches` in total for season 2022.\

#### Which teams have reversed their matches?
Now, it is time to know which teams accomplish a reverse sweep in this season and against who.  
```{r achieve the reversed sweep, echo=FALSE}
owl_reversed <- owl_data_2022_scores %>%
  select(match_id, game_number, stage_class, map_winner, match_winner) %>% 
  group_by(match_id, game_number) %>% 
  filter(row_number() == 1, stage_class != "Postseason") %>%
  group_by(match_id) %>% 
  filter(game_number <= 2) %>%
  mutate(winner_second_map = lag(map_winner)) %>% 
  filter(game_number == 2) %>% 
  mutate(reversed_possible = ifelse(map_winner == winner_second_map, 1, 0),
         reversed = ifelse(reversed_possible == 1 & match_winner != map_winner,
                           "Reversed", "No")) %>% 
  ungroup() %>% 
  filter(reversed == "Reversed") %>% 
  select(-reversed_possible, -reversed, -stage_class, -game_number, -map_winner, -match_id) %>% 
  rename("match_loser" = winner_second_map)

owl_reversed %>% 
  group_by(match_winner) %>% 
  count() %>%
  left_join(owl_images, c("match_winner" = "team_name")) %>% 
  ggplot(aes(x = n, y = reorder(match_winner, n))) +
  geom_image(aes(image = team_image), size = 0.05) +
  labs(y = NULL,
       x = "Number of reverse sweep accomplished",
       title = "Who accomplished a reverse sweep?",
       caption = "Source: https://overwatchleague.com") +
  scale_x_continuous(labels = c(0:4),
                     breaks = c(0:4),
                     expand = expansion(add = 1))
```

#### Who got reversed sweep?
```{r got the reversed sweep, echo=FALSE}
owl_reversed %>% 
  group_by(match_winner) %>% 
  mutate(num_reverse = n()) %>% 
  group_by(match_winner, match_loser) %>%
  mutate(num_match = row_number(),
         num_match = max(num_match)) %>% 
  left_join(owl_images, c("match_loser" = "team_name")) %>% 
  ggplot(aes(x = match_loser, y = reorder(match_winner, num_reverse))) +
  geom_image(aes(image = team_image), size = 0.05) +
  geom_text(aes(label = num_match), nudge_x = 0.5) +
  labs(y = "Winner team",
       x = "Number of reverse sweep accomplished",
       title = "Who accomplished a reverse sweep?",
       subtitle = "Against who? How many times?",
       caption = "Source: https://overwatchleague.com") +
  scale_x_discrete(labels = NULL)
```

### Time to win TODO

The average time to win and the number of matches can be helpful to know things like:\
- *How constant is a team at wining*.\
- The *time window* that a particular team has to win, longer times might impact certain teams.\
- The play style of teams, like *aggro, stall or control*.\
```{r time to win, echo=FALSE, message=FALSE}
owl_data_2022_scores_wins <- owl_data_2022_scores %>% 
  select(match_id, game_number, stage_class, map_winner, map_loser, team_region,
         team_home, total_time_match, match_winner) %>% 
  group_by(match_id) %>% 
  filter(row_number() == 1, stage_class %in% "Qualifiers")

owl_data_2022_scores_wins %>% 
  ungroup() %>% 
  summarise(total_time = sum(total_time_match),
            total_wins = sum(game_number),
            time_to_win = total_time / total_wins,
            time_to_win = as.numeric(as.duration(time_to_win), "hour"))

owl_data_2022_scores_wins %>% 
  group_by(team_region, match_winner) %>% 
  summarise(total_time = sum(total_time_match),
            total_wins = sum(game_number),
            time_to_win = total_time / total_wins,
            time_to_win = as.numeric(as.duration(time_to_win), "hour")) %>% 
  arrange(team_region, time_to_win, total_wins, total_time) %>% 
  left_join(owl_images, c("match_winner" = "team_name")) %>% 
  ggplot(aes(x = time_to_win,
             y = reorder(match_winner, desc(time_to_win)))) +
  geom_image(aes(image = team_image)) +
  geom_vline(xintercept = 1.23, colour = "red") +
  geom_text(aes(x = 1.3, y = 20,label = "League average", family = "serif"),
            colour = "red") +
  labs(x = "Average time of the match (hours)",
       y = NULL,
       title = "Average time to achive a win.",
       caption = "Source: https://overwatchleague.com") +
  scale_x_continuous(expand = expansion(0.5))

```

#### Win vs Time
Talking a little bit about of the time to get a win and the total matches won.\
> In the future I will try to do a shiny app for interactivity, I have the function for it, but no the knowledge to make the app.

```{r Correlation, echo=FALSE, message=FALSE}
win_base <- owl_data_2022_scores_wins %>% 
  group_by(team_region, match_winner) %>% 
  summarise(total_time = sum(total_time_match),
            total_wins = sum(game_number),
            time_to_win = total_time / total_wins,
            time_to_win = as.numeric(as.duration(time_to_win), "hour")) %>% 
  arrange(team_region, time_to_win, total_wins, total_time) 

win_base %>% 
  left_join(owl_images, c("match_winner" = "team_name")) %>% 
  ggplot(aes(x = total_wins,y = time_to_win)) +
  geom_image(aes(image = team_image)) +
  labs(x = "Matches wins",
       y = "Average time of the match (hours)",
       colour = "Number of wins",
       title = "Correlation between wins and the time to win.",
       caption = "Source: https://overwatchleague.com") 
```

### What's next?
I haven't finish the whole analysis, I need to develop the charts for some topics that I already have like: `global win rate`, `map winrate by team`, `First map win the match`.\
Also look into `distances, progress for the type of maps`, `individual stats` that are in other dataset, `etc`.\
Finally, continue **improving the analysis and visualisations**, and look into the **following season**.
