---
title: "Overwatch League Season 2022"
author: "Israel Garcia C."
date: "`r Sys.Date()`"
output: 
 html_document:
  theme:
    bootswatch: darkly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(DT)
library(lubridate)
library(tidyverse)
owl_data_2022_scores <- read_rds("Data_all\\owl_data_2022_scores.rds")
```

## Overwatch League Season 2022 Analysis
Thanks to the [official page](https://overwatchleague.com/en-us/statslab) of the league, people can search and play with the stats
produced in each match. Although it provides an easy way to get insight, I found it difficult to use for some of the insights and
hypothesis.  
Also, it includes information of previous seasons, but they have a **different format** both in the tournaments and in the
number of players, **different map** pools, between other aspects that makes the analysis of the current season with previous information
kind of useless.  
Some general considerations:  
1. The last update of the official dataset was in `10/26/2022`, but the season has already finish.\
1.1 This makes that the analysis incomplete, missing part of the information for the `Postseason` and all information of the `Playoffs`.\
2. This version might not be as interactive as in the official page, I plan to make it a `Shiny App` later or maybe for the following season. Another option was to make it using `SQL` and `Power BI/Tableau`, but for practice I choose to make it all in R.\
3. Graphics need a correct theme, I know but I am doing this in my free time as data analyst. I am trying to improve analysis and visualsation.

### Preview
As always, it is useful for the **user** and the developer to have a preview of the data in the analysis.  
Although the data in the official dataset is *almost* clean, there are some data entry errors and missing useful columns.  
Here, you can see a clean preview, first 5 entries, with additional information that I have added.
```{r Preview, echo = FALSE, results='asis'}
datatable(head(owl_data_2022_scores, 20), 
          options = list(
          columnDefs = list(list(className = 'dt-center', targets = 5)),
          pageLength = 5,
          lengthMenu = c(5, 10, 15, 20),
          extensions = 'FixedColumns',
          dom = 't',
          scrollX = TRUE,
          fixedColumns = TRUE))
```


### Match duration
The first part is to look into the match duration, in hours, for all the matches.  
As expected, *Tournaments* tend to have a larger duration than Qualifiers, this can be explained by factors such as when you lose in a tournament you are knockoff, and that finals matches are *first to 4* and not *first to 3*.
```{r Distribution of the matches duration in Hours, echo=FALSE, warning = FALSE}
owl_data_2022_scores %>% 
  select(match_id, game_number, stage_class, total_time_match) %>% 
  group_by(match_id) %>% 
  filter(row_number() == 1, stage_class != "Postseason") %>%
  ggplot(aes(x = as.numeric(as.duration(total_time_match), "hour"))) +
  geom_histogram(binwidth = 0.1) +
  facet_grid(stage_class ~ ., scales = "free_y") + 
  labs(x = "Total match time (Hours)",
       y = "Total matches",
       title = "OWL Season 2022 Match duration",
       caption = "Source: https://overwatchleague.com")

owl_data_2022_scores %>%  
  select(match_id, game_number, match_rounds, stage_class, total_time_match) %>% 
  group_by(match_id) %>% 
  filter(game_number == max(game_number), match_rounds == max(match_rounds), stage_class != "Postseason") %>%
  ggplot(aes(x = as.numeric(as.duration(total_time_match), "hour"), y = match_rounds)) +
  geom_boxplot() +
  coord_flip() +
  facet_grid(stage_class ~ ., scales = "free_y") +
  labs(x = "Total match time (Hours)",
       y = "Total rounds per match",
       title = "OWL Season 2022 Match duration",
       caption = "Source: https://overwatchleague.com")
```

### Map pool
Overwatch counts with a variety of maps, some of them are eligible during certain qualifiers or tournaments.  
In general, the first team to win 3 maps win the match. The order of the maps according to their type is:   
1. Control map  
2. Escort map  
3. Hybrid map, at this point it is *expected* that a team has won by a score of *3-0*, only in Qualifiers.  
4. Push map  
5. Another control map  
In the following figure, we can see that Escort and Hybrid maps have the same number of maps played as they are not repeated in case of ties. Push has the lowest as is not as common to see a fourth map, finally Control has the most maps played because is functions as a tie breaker as it is impossible to tie in that type of map. 
```{r Map pool, echo=FALSE, warning = FALSE}
owl_data_2022_scores %>%
  select(match_id, game_number, match_winner, map_name, total_time_match, stage_class, map_class) %>% 
  group_by(match_id, game_number) %>% 
  filter(row_number() == 1, stage_class != "Postseason") %>% 
  ggplot(aes(x = map_class, fill = map_class)) +
  geom_bar() +
  coord_flip() +
  facet_grid(stage_class ~ .) +
  labs(x = NULL,
       y = "Total times played",
       title = "OWL Season 2022",
       subtitle = "Maps played during this season",
       caption = "Source: https://overwatchleague.com",
       fill = "Map type") +
  paletteer::scale_fill_paletteer_d("rtist::vangogh")
```

### Scores in Overwatch League
As mentioned previously, Qualifiers is played in a format *first to 3*, while Tournaments is *first to 4*. 
```{r Common scores, echo=FALSE}
owl_data_2022_scores_common <- owl_data_2022_scores %>% 
  select(match_id, score_winner, score_loser, stage_class) %>% 
  group_by(match_id) %>% 
  filter(row_number() == 1) %>%
  mutate(score_str = str_c(score_winner, score_loser, sep = "-"))

owl_data_2022_scores_common %>%
  filter(stage_class != "Postseason") %>%
  ggplot(aes(x = score_str, fill = score_str)) +
  geom_bar() +
  coord_flip() +
  facet_wrap(. ~ stage_class, nrow = 2) +
  labs(x = NULL,
       y = "Total matches",
       title = "Common scores during season 2022",
       caption = "Source: https://overwatchleague.com",
       fill = "Score") +
  paletteer::scale_fill_paletteer_d("colorBlindness::paletteMartin")
```

### Maps taken
As in any sports, both teams fight to get points for the final score. In Overwatch each point is attributed by winning or losing a map. Obviously, the number of maps taken by losers will be less than the winners.
```{r Maps taken, echo = FALSE, message=FALSE}
owl_data_2022_scores %>%
  select(match_id, score_winner, score_loser) %>% 
  group_by(match_id) %>% 
  filter(row_number() == 1) %>%
  ungroup() %>% 
  summarise(`Maps taken by winning team` = sum(score_winner),
            `Maps taken by loser team` = sum(score_loser))

owl_data_2022_scores_maps_taken <- owl_data_2022_scores %>% 
  select(match_id, game_number, map_class, map_winner, match_winner, stage_class) %>% 
  group_by(match_id, game_number) %>% 
  filter(row_number() == 1)

owl_data_2022_scores_maps_taken %>% 
  group_by(map_class) %>% 
  summarise(wins = ifelse((map_winner == match_winner), 1 , 0)) %>%
  mutate(wins = ifelse(wins == 1,"Winner", "Loser")) %>% 
  count(wins) %>% 
  ggplot(aes(x = map_class, y = n)) +
  geom_bar(aes(fill = wins), stat = "identity", position = "dodge") +
  labs(x = NULL,
       y = "Total maps",
       title = "Maps taken by each team",
       caption = "Source: https://overwatchleague.com",
       fill = "Map taken by:") +
  scale_fill_brewer(palette = "Set1")

owl_data_2022_scores_maps_taken %>% 
  group_by(map_class, stage_class) %>% 
  filter(stage_class != "Postseason") %>% 
  summarise(wins = ifelse((map_winner == match_winner), 1 , 0)) %>%
  count(wins) %>% 
  mutate(wins = ifelse(wins == 1,"Winner", "Loser")) %>% 
  ggplot(aes(x = map_class, y = n)) +
  geom_bar(aes(fill = wins), stat = "identity", position = "dodge") +
  coord_flip() + 
  facet_wrap(. ~ stage_class, nrow = 2) +
  labs(x = NULL,
       y = "Total maps",
       title = "Maps taken by the team",
       caption = "Source: https://overwatchleague.com",
       fill = "Map taken by:") +
  scale_fill_brewer(palette = "Set1")
```

### Reverse sweep
One of the most entertain/frustrating moments in any sports are the reverse sweep.\
In Overwatch a reverse sweep happens when a team won the first 2 maps, but lose the following 3 maps in a row.\
```{r reversed sweep, echo=FALSE}
owl_data_2022_scores %>%
  select(match_id, game_number, stage_class, map_winner, match_winner) %>% 
  group_by(match_id, game_number) %>% 
  filter(row_number() == 1, stage_class != "Postseason") %>%
  group_by(match_id) %>% 
  filter(game_number <= 2) %>%
  mutate(winner_second_map = lag(map_winner)) %>% 
  filter(game_number == 2) %>% 
  mutate(reversed_possible = ifelse(map_winner == winner_second_map, 1, 0),
         reversed = ifelse(reversed_possible == 1 & match_winner != map_winner,
                           "Reversed", "No")) %>% 
  ungroup() %>% 
  count(reversed)
```

From the numbers, it seems that a reverse sweep is quite infrequent, just `18 matches` in total for season 2022.\
Now, is time to know the following:\

#### Which teams have reversed their matches?
```{r achieve the reversed sweep, echo=FALSE}
owl_data_2022_scores %>%
  select(match_id, game_number, stage_class, map_winner, match_winner) %>% 
  group_by(match_id, game_number) %>% 
  filter(row_number() == 1, stage_class != "Postseason") %>%
  group_by(match_id) %>% 
  filter(game_number <= 2) %>%
  mutate(winner_second_map = lag(map_winner)) %>% 
  filter(game_number == 2) %>% 
  mutate(reversed_possible = ifelse(map_winner == winner_second_map, 1, 0),
         reversed = ifelse(reversed_possible == 1 & match_winner != map_winner,
                           "Reversed", "No")) %>% 
  ungroup() %>% 
  filter(reversed == "Reversed") %>% 
  group_by(match_winner) %>% 
  count() %>%
  ggplot(aes(x = n, y = reorder(match_winner, n))) +
  geom_point(size = 2) +
  labs(y = NULL,
       x = "Number of matches",
       title = "Who accomplished the reverse sweep?",
       caption = "Source: https://overwatchleague.com") +
  scale_x_continuous(labels = c(1:4),
                     breaks = c(1:4),
                     minor_breaks = NULL)
```

#### Who got reversed sweep?
```{r got the reversed sweep, echo=FALSE}
owl_data_2022_scores %>%
  select(match_id, game_number, stage_class, map_winner, match_winner) %>% 
  group_by(match_id, game_number) %>% 
  filter(row_number() == 1, stage_class != "Postseason") %>%
  group_by(match_id) %>% 
  filter(game_number <= 2) %>%
  mutate(winner_second_map = lag(map_winner)) %>% 
  filter(game_number == 2) %>% 
  mutate(reversed_possible = ifelse(map_winner == winner_second_map, 1, 0),
         reversed = ifelse(reversed_possible == 1 & match_winner != map_winner,
                           "Reversed", "No")) %>% 
  ungroup() %>% 
  filter(reversed == "Reversed") %>% 
  group_by(map_winner) %>% 
  count() %>%
  ggplot(aes(x = n, y = reorder(map_winner, n))) +
  geom_point(size = 2) +
  labs(y = NULL,
       x = "Number of matches",
       title = "Who got reverse sweep?",
       caption = "Source: https://overwatchleague.com") +
  scale_x_continuous(labels = c(1:3),
                     breaks = c(1:3),
                     minor_breaks = NULL)
```

### Time to win
The average time to win and the number of matches can be helpful to know things like:\
- *How constant is a team at wining*.\
- The *time window* that a particular team has to win, longer times might impact certain teams.\
- The play style of teams, like *aggro, stall or control*.\
```{r time to win, echo=FALSE, message=FALSE}
owl_data_2022_scores_wins <- owl_data_2022_scores %>% 
  select(match_id, game_number, stage_class, map_winner, map_loser, team_region,
         team_home, total_time_match, match_winner) %>% 
  group_by(match_id) %>% 
  filter(row_number() == 1, stage_class %in% "Qualifiers")

owl_data_2022_scores_wins_table <- owl_data_2022_scores_wins %>% 
  group_by(team_region, match_winner) %>% 
  summarise(total_time = sum(total_time_match),
            total_wins = sum(game_number),
            time_to_win = total_time / total_wins,
            time_to_win = as.numeric(as.duration(time_to_win), "hour"),
            time_to_win = round(time_to_win, 2)) %>%
  select(team_region, match_winner, time_to_win, total_wins) %>% 
  arrange(team_region, match_winner, time_to_win, total_wins) %>% 
  rename("team region" = team_region,
         "team" = match_winner,
         "average time to get a win" = time_to_win,
         "total match wins" = total_wins)

datatable(owl_data_2022_scores_wins_table, 
          options = list(
          pageLength = 5))

owl_data_2022_scores_wins %>% 
  group_by(team_region, match_winner) %>% 
  summarise(total_time = sum(total_time_match),
            total_wins = sum(game_number),
            time_to_win = total_time / total_wins,
            time_to_win = as.numeric(as.duration(time_to_win), "hour")) %>% 
  arrange(team_region, time_to_win, total_wins, total_time) %>% 
  ggplot(aes(x = time_to_win,
             y = reorder(match_winner, desc(time_to_win)))) +
  geom_point() +
  geom_vline(xintercept = 1.23, colour = "red") +
  geom_text(aes(x = 1.3, y = 20,label = "League average", family = "serif"),
            colour = "red") +
  labs(x = "Average time of the match (hours)",
       y = NULL,
       title = "Average time to achive a win.",
       caption = "Source: https://overwatchleague.com")
```

#### Win vs Time
Talking a little bit about of the time to get a win and the total matches won.\
> In the future I will try to do a shiny app for interactivity, I have the function for it, but no the knowledge to make the app.

```{r Correlation, echo=FALSE, message=FALSE}
win_base <- owl_data_2022_scores_wins %>% 
  group_by(team_region, match_winner) %>% 
  summarise(total_time = sum(total_time_match),
            total_wins = sum(game_number),
            time_to_win = total_time / total_wins,
            time_to_win = as.numeric(as.duration(time_to_win), "hour")) %>% 
  arrange(team_region, time_to_win, total_wins, total_time) 

win_base %>% 
  ggplot(aes(x = total_wins,y = time_to_win, colour = total_wins)) +
  geom_point(size = 2) +
  geom_point(data = filter(win_base, match_winner== "Seoul Dynasty"),
             colour = "#DAA520",
             size = 3) +
  annotate(geom = "text", x = 17, y = 1.13, label = "Seoul Dynasty") + 
  labs(x = "Matches wins",
       y = "Average time of the match (hours)",
       colour = "Number of wins",
       title = "Correlation between wins and the time to win.",
       caption = "Source: https://overwatchleague.com") 
```

### What's next?
I haven't finish the whole analysis, I need to develop the charts for some topics that I already have like: `global win rate`, `map winrate by team`, `First map win the match`.\
Also look into `distances, progress for the type of maps`, `individual stats` that are in other dataset, `etc`.\
Finally, continue **improving the analysis and visualisations**, and look into the **following season**.
